---
title: "Attention Competition Model"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Attention Competition Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Theory background

Attention can be understood as a selection process.

Signals may compete based on salience, novelty, goal relevance, and noise.

## Run the model

```{r attention-model}
attn <- attention_competition_model(n_signals = 6, steps = 100, weights = c(0.4, 0.3, 0.3), seed = 22)
head(attn)
```

## Which signals were selected?

```{r selected-signals}
table(attn$selected_signal)
```

## Plot priorities

```{r priority-plot}
plot_consciousness_sim(attn, x = "step", y = "priority", group = "signal")
```

## Interpretation

Changing weights changes the basis of competition. A salience-dominated system responds strongly to intense signals, while a goal-dominated system prioritizes task relevance.
