---
title: "Broadcast Network Simulation"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Broadcast Network Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Theory background

In global workspace models, a selected signal becomes widely available to multiple systems.

`broadcast_network()` represents this idea using a simple network.

## Run the model

```{r broadcast-model}
net <- broadcast_network(n_nodes = 12, steps = 60, source_node = 1, connection_probability = 0.25, seed = 7)
head(net$time_series)
```

## View network structure

```{r adjacency}
net$adjacency_matrix
```

## Plot activation spread

```{r broadcast-plot}
plot_consciousness_sim(net$time_series, x = "step", y = "activation", group = "node")
```

## Interpretation

More connected networks tend to spread activation more broadly. Sparse networks may isolate activation in a small part of the system.
