---
title: "Comparing Consciousness Theories"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Comparing Consciousness Theories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(consciousnessModelR)
```


## Purpose

This article compares the theories represented in `consciousnessModelR`. The goal is not to rank theories, but to clarify what each theory emphasizes and how the package translates those ideas into educational simulations.

## Comparison table

| Theoretical emphasis | Package function | Main modeling idea |
|---|---|---|
| Global access | `simulate_global_workspace()` | Processes compete; a winner may ignite and become globally available. |
| Attention | `attention_competition_model()` | Signals compete based on salience, novelty, and goal relevance. |
| Broadcast | `broadcast_network()` | A selected signal spreads through a network. |
| Integration | `simulate_information_integration()` | Components combine connectivity, shared activity, and differentiation. |
| Threshold crossing | `consciousness_threshold()` | Activation is classified relative to a simplified access threshold. |

## A compact workflow

```{r compare-workflow}
gw <- simulate_global_workspace(seed = 11)

thresholded <- consciousness_threshold(
  gw,
  activation_col = "activation",
  threshold = 0.70
)

head(thresholded)
```

## Attention before access

```{r compare-attention}
attn <- attention_competition_model(seed = 11)

table(attn$selected_signal)
```

## Broadcast after selection

```{r compare-broadcast}
net <- broadcast_network(seed = 11)

head(net$time_series)
```

## Integration as a different explanatory level

```{r compare-integration}
info <- simulate_information_integration(seed = 11)

info$summary
```

## Interpretation

The functions represent different explanatory levels.

Attention models ask which signal is prioritized. Global workspace models ask whether selected information becomes globally available. Broadcast models ask how information spreads. Integration models ask whether the system has unified but differentiated internal structure. Threshold models ask when a continuous process is classified as access-like.

These are not necessarily mutually exclusive. A complete theory of consciousness may require several levels of explanation.

## Reflection questions

1. Can a signal be attended without being conscious?
2. Can information be globally broadcast without rich subjective experience?
3. Can a system be integrated but unable to report?
4. Does threshold crossing explain consciousness or only access?
5. What would a stronger model need to include?

## Suggested readings

- Baars, B. J. (1988). *A Cognitive Theory of Consciousness.*
- Chalmers, D. J. (1996). *The Conscious Mind.*
- Dehaene, S. (2014). *Consciousness and the Brain.*
- Seth, A. K. (2018). Consciousness: The last 50 years and the next.
- Tononi, G. (2004). An information integration theory of consciousness.
