---
title: "Threshold Models of Awareness-like Processing"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Threshold Models of Awareness-like Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```

## Theory background

Some theories describe conscious access as threshold-like.

A stimulus may remain subliminal if activation is weak, but become reportable or globally available if activation crosses a threshold.

## Run and apply threshold

```{r threshold-sim}
sim <- simulate_global_workspace(seed = 3, ignition_threshold = 0.75)
thresholded <- consciousness_threshold(sim, activation_col = "activation", threshold = 0.70)
head(thresholded)
```

## Count threshold crossings

```{r threshold-count}
table(thresholded$above_threshold)
```

## Interpretation

Rows above the threshold represent signals that meet a simplified criterion for awareness-like processing. This is not a claim about subjective experience.
