---
title: "Getting Started with consciousnessModelR"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Getting Started with consciousnessModelR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Overview

`consciousnessModelR` provides simplified simulations for teaching major ideas in consciousness studies.

The package focuses on educational clarity. It does not attempt to measure or produce consciousness.

The central learning question is:

> How can different theories of consciousness be represented as simple computational models?

## Main ideas

The package includes models inspired by:

- Global Workspace Theory
- Attention competition
- Broadcast dynamics
- Information integration
- Threshold models of awareness-like processing

## A first simulation

```{r first-simulation}
sim <- simulate_global_workspace(n_processes = 8, steps = 100, seed = 123)
head(sim)
```

## Plot activation through time

```{r first-plot}
plot_consciousness_sim(sim, x = "step", y = "activation", group = "process")
```

## Apply a threshold

```{r first-threshold}
thresholded <- consciousness_threshold(sim, activation_col = "activation", threshold = 0.7)
head(thresholded)
```

## Interpretation

In this toy model, several processes compete for access. A process with sufficiently high activation may cross a threshold and become globally broadcast.

This is an educational representation of one idea from Global Workspace Theory, not a model of real subjective experience.

## Suggested reading

Global Workspace Theory was developed by Baars and later expanded in cognitive neuroscience by Dehaene and colleagues [@baars1988cognitive; @dehaene2011experimental; @dehaene2014consciousness].
