---
title: "Global Workspace Simulation"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Global Workspace Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Theory background

Global Workspace Theory proposes that conscious access involves information becoming globally available to many specialized systems.

In simple terms:

1. Many processes operate locally.
2. They compete for attention.
3. One signal may become dominant.
4. The winning signal is broadcast to the wider system.

## Run the simulation

```{r gw-simulation}
gw <- simulate_global_workspace(n_processes = 10, steps = 120, noise = 0.1, ignition_threshold = 0.75, seed = 10)
head(gw)
```

## Visualize activation

```{r gw-plot}
plot_consciousness_sim(gw, x = "step", y = "activation", group = "process")
```

## Inspect winning processes

```{r gw-winners}
table(gw$winner)
```

## Interpretation

A process is treated as entering the workspace when it has the highest activation and crosses the ignition threshold.

Low thresholds make broadcast easier. High thresholds make broadcast rarer.
