---
title: "Information Integration Simulation"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Information Integration Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Theory background

Integrated Information Theory argues that consciousness is related to how much information is integrated within a system.

This package does not compute IIT's formal phi. Instead, it creates an educational proxy based on connectivity, shared activity, and differentiation.

## Run the model

```{r integration-model}
info <- simulate_information_integration(n_components = 8, steps = 100, connection_probability = 0.3, seed = 42)
info$summary
```

## Plot component activity

```{r integration-plot}
plot_consciousness_sim(info$time_series, x = "step", y = "activation", group = "component")
```

## Compare sparse and dense systems

```{r integration-compare}
sparse <- simulate_information_integration(connection_probability = 0.10, seed = 42)
dense <- simulate_information_integration(connection_probability = 0.70, seed = 42)
sparse$summary
dense$summary
```

## Limitation

This function is not an implementation of Integrated Information Theory and does not compute phi [@tononi2004integrated; @oizumi2014phenomenology].
