---
title: "Information Integration"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Information Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(consciousnessModelR)
```


## Purpose

This article explains the theoretical background behind `simulate_information_integration()`. The function is inspired by information-integration approaches to consciousness, especially the idea that consciousness may depend on both integration and differentiation.

The function does not calculate formal phi and should not be presented as an implementation of Integrated Information Theory.

## Integration and differentiation

A system can be highly differentiated if its parts carry distinct patterns of activity. A system can be highly integrated if its parts influence one another and behave as a coordinated whole.

A common intuition in information-integration theories is that consciousness requires both. A system that is completely fragmented lacks unity. A system that is completely uniform lacks differentiated content.

## Relation to the package

`simulate_information_integration()` creates a simple network of components. Components influence one another through connections. The function then computes an educational integration score based on average connectivity, shared activity among components, and differentiation among component states.

This score is not a clinical, neuroscientific, or philosophical measure of consciousness.

## Basic simulation

```{r integration-basic}
info <- simulate_information_integration(
  n_components = 8,
  steps = 100,
  connection_probability = 0.30,
  seed = 42
)

info$summary
```

```{r integration-plot}
plot_consciousness_sim(
  info$time_series,
  x = "step",
  y = "activation",
  group = "component"
)
```

## Comparing sparse and dense systems

```{r integration-comparison}
sparse <- simulate_information_integration(
  connection_probability = 0.10,
  seed = 42
)

moderate <- simulate_information_integration(
  connection_probability = 0.30,
  seed = 42
)

dense <- simulate_information_integration(
  connection_probability = 0.70,
  seed = 42
)

rbind(
  sparse = sparse$summary,
  moderate = moderate$summary,
  dense = dense$summary
)
```

## Interpretation

Sparse systems may show weak integration because components are not strongly connected. Very dense systems may show high coupling but reduced differentiation if all components behave similarly. Moderate systems may sometimes balance integration and differentiation.

This illustrates a central theoretical tension: a system must be unified enough to function as a whole, but differentiated enough to support rich content.

## Relation to other functions

This model complements `broadcast_network()`. Broadcast models emphasize access and availability. Integration models emphasize internal structure and causal organization.

## Limitations

This simulation is only an educational proxy. Formal Integrated Information Theory is mathematically complex and has specific definitions that are not implemented here.

## Suggested readings

- Tononi, G. (2004). An information integration theory of consciousness.
- Oizumi, M., Albantakis, L., & Tononi, G. (2014). Integrated Information Theory 3.0.
- Seth, A. K., Barrett, A. B., & Barnett, L. (2011). Causal density and integrated information.
- Chalmers, D. J. (2016). The combination problem for panpsychism.
