---
title: "Limitations and Responsible Use"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Limitations and Responsible Use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r load-package}
library(consciousnessModelR)
```


## Why this matters

Computational models of consciousness can be misleading if presented too strongly.

The functions in `consciousnessModelR` are toy educational models. They are designed to help learners compare theory structures, not to evaluate real minds.

## What the package does not do

The package does not:

- Detect consciousness
- Measure subjective experience
- Diagnose cognitive states
- Determine whether AI systems are conscious
- Represent detailed brain physiology
- Resolve the hard problem of consciousness

## Responsible interpretation

A simulation may show competition, broadcast, integration, threshold crossing, or network activation. These are formal patterns, not proof of conscious experience.

## Educational value

The models are useful because they make abstract ideas concrete.

Students can ask:

- What assumptions does each theory make?
- What variables matter?
- What outputs would count as evidence?
- What is missing from the model?
- How could the model be improved?

## Suggested readings

For broader philosophical and scientific context, see Chalmers, Seth, Baars, Dehaene, and Tononi [@chalmers1996conscious; @seth2018consciousness; @baars1988cognitive; @dehaene2014consciousness; @tononi2004integrated].
